---
title: "Course Project 1"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, echo=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)
library(xtable)
```

## 1. Code for reading in the dataset and/or processing the data

```{r read, results='asis'}
if (!file.exists("./data/getdata4.zip")){
  fileurl<-"https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
  download.file(fileurl,"./data/getdata4",method="curl")
}
if (!(file.exists("./data/activity.csv"))) { 
  unzip("./data/getdata3.zip",exdir="./data") 
}
data<-read.csv("./data/activity.csv")
xt<-xtable(head(data))
print(xt,type="html")
```

## 2. Histogram of the total number of steps taken each day

```{r histogram, results='asis'}
data_hist<-data%>%group_by(date)%>%summarise(steps=sum(steps))%>%filter(!is.na(steps))
xt<-xtable(head(data_hist))
print(xt,type="html")
hist(data_hist$steps,xlab = "Steps taken each day",main = NULL)
```

## 3. Mean and median number of steps taken each day

```{r mean}
mean(data_hist$steps)
median(data_hist$steps)
```

## 4. Time series plot of the average number of steps taken

```{r timeseries, results='asis'}
data_timeseries<-data%>%select(interval,steps)%>%filter(!is.na(steps))%>%group_by(interval)%>%summarise(steps=mean(steps))
xt<-xtable(head(data_timeseries))
print(xt,type="html")
ggplot(data_timeseries,aes(interval,steps))+geom_line()
```

## 5. The 5-minute interval that, on average, contains the maximum number of steps

```{r maxsteps}
data_timeseries$interval[which.max(data_timeseries$steps)]
```

## 6. Code to describe and show a strategy for imputing missing data
### Missing values populated by the mean of steps taken per day
```{r imputemissing, results='asis'}
sprintf("Missing values = %1.2f%%", 100*mean(is.na(data)))
impute_missing<-function(x){replace(x,is.na(x),mean(x,na.rm = TRUE))}
data_filled<-data%>%group_by(date)%>%summarise(steps=sum(steps))%>%mutate(steps=impute_missing(steps))
xt<-xtable(head(data_filled))
print(xt,type="html")
```

## 7. Histogram of the total number of steps taken each day after missing values are imputed
```{r histogram_missing}
hist(data_filled$steps,xlab = "Steps taken each day",main = NULL)
sprintf("Old mean =%.1f and New mean = %.1f",mean(data_hist$steps),mean(data_filled$steps))
sprintf("Old median =%.1f and New median = %.1f",median(data_hist$steps),median(data_filled$steps))
```

## 8. Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends
```{r panelplot,results='asis'}
data_day<-data%>%group_by(interval)%>%mutate(steps=impute_missing(steps))%>%mutate(date=as.Date(date))%>%mutate(day=weekdays(date))%>%mutate(weekend=ifelse(day=="Saturday"|day=="Sunday", "Weekend", "Weekday" ))
data_day<-aggregate(data_day$steps,by=list(data_day$weekend,data_day$interval),na.omit(mean))
names(data_day) <- c("weekend", "interval", "steps")
xt<-xtable(head(data_day))
print(xt,type="html")
ggplot(data_day, aes(interval,steps, color=weekend)) + geom_line()+facet_grid(weekend ~.)+xlab("Interval")+ylab("Mean of Steps") +ggtitle("Comparison of Average Number of Steps per 5-minute interval")+theme(legend.position = "none")
```