---
title: "Statistical Inference Project - Part 2"
author: "Srotaswati Panda"
date: "01/10/2019"
output: pdf_document
geometry: margin=1.5cm
---

### This is a project for the Coursera Data Science class on Statistical Inference

# Part 1: Basic Inferential Data Analysis
### Overview: 
Analyze the ToothGrowth data in the R datasets package.  
1. Load the ToothGrowth data and perform some basic exploratory data analyses.  
2. Provide a basic summary of the data.  
3. Use confidence intervals and/or hypothesis tests to compare tooth growth by supp and dose. (Only use the techniques from class, even if there's other approaches worth considering)  
4. State your conclusions and the assumptions needed for your conclusions.

### Setup Environment and Load Data
Load libraries and set global options
```{r setup}
library(knitr)
opts_chunk$set(echo = TRUE,message=FALSE,warning=FALSE,tidy=TRUE)
library(ggplot2)
library(datasets)
library(xtable)
```

### 1. Load ToothGrowth dataset and provide basic summary of the data
Display the first few rows of the data and show the unique values of Dose and Supplement Delivery Method
```{r dataset, results='asis'}
data(ToothGrowth)
print(kable(head(ToothGrowth)))
print(kable(table(ToothGrowth$supp,ToothGrowth$dose)))
```
```{r summary}
sum(!complete.cases(ToothGrowth))
summary(ToothGrowth$len)
```

### 2. Perform basic exploratory analyses
Boxplots would show the range of Tooth Lengths vis a vis Dose Amount and Supplement Delivery Method. Tooth Lengths seem to be positively correlated with both Dose and Delivery Method. Further evidence can be established by running t-tests.
```{r boxplot}
ToothGrowth$dose<-as.factor(ToothGrowth$dose)
g<-ggplot(data=ToothGrowth,aes(x=dose,y=len))
g<-g+geom_boxplot(aes(fill=dose))
g<-g+labs(x="Dose Amount",y="Tooth Length")
g<-g+facet_grid(.~supp)
g
g<-ggplot(data=ToothGrowth,aes(x=supp,y=len))
g<-g+geom_boxplot(aes(fill=supp))
g<-g+labs(x="Supplement Delivery",y="Tooth Length")
g<-g+facet_grid(.~dose)
g
```

### 3. Hypothesis testing
```{r ttests1}
t.test(len~supp,data = ToothGrowth)
```
At the significance level ${\alpha}$ = 0.5, the p-Value is higher at 0.06, we fail to reject the null hypothesis ${H_o}$: difference in means for the groups 'OJ' and 'VC' is equal to 0. Also, the confidence interval includes 0, so the test is not significant.

```{r ttests2}
sub<-subset(ToothGrowth,ToothGrowth$dose %in% c(0.5,1.0))
t.test(len~dose,data = sub)
sub<-subset(ToothGrowth,ToothGrowth$dose %in% c(1.0,2.0))
t.test(len~dose,data = sub)
sub<-subset(ToothGrowth,ToothGrowth$dose %in% c(2.0,0.5))
t.test(len~dose,data = sub)
```
At the significance level ${\alpha}$ = 0.5, the p-Values of all the above tests is very close to 0 which implies that the null hypothesis ${H_o}$ will be rejected in favor of the alternative hypothesis ${H_a}$ that the difference in means for the three groups of Dosage Amounts in not equal to 0. Also, the confidence intervals do not contain 0.

### 4. Conclusions
It is not very clear if the Tooth Lengths are independent of Delivery Methods. However from the t-tests and the confidence intervals, Tooth Lengths increase with the increase in Dosage Amounts, which is also evident from the boxplots.